\textbf{Evaluation of machine learning methods for 16S rRNA gene data}

### 2019-05-15 submission/manuscript.tex
manuscript update with fig 3

Advances in sequencing technology and decreasing costs of generating 16S
rRNA gene sequences have allowed rapid exploration of human associated
microbiome and its health implications. Currently, the human microbiome
field is growing at an unprecedented rate and as a result, there is an
increasing demand for methods that identify associations between members
of the microbiome and human health. However, this is difficult as human
associated microbial communities are remarkably complex and uneven. It
is unlikely that a single species can explain a disease. Instead,
subsets of those communities, in relation to one another and to their
host, account for the differences in health outcomes. Machine learning
(ML) methods are effective at recognizing and highlighting patterns in
complex microbial datasets. Therefore, researchers have started to
explore the utility of ML models that use microbiota associated
biomarkers to predict human health and to understand the microbial
ecology of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases (IBD), obesity, and type 2 diabetes (1--11).
However, currently the field's use of ML lacks clarity and consistency
on which methods are used and how these methods are implemented. Most
notably, there are misleading practices of using ``leaky'' ML pipelines
where models are tested on training data or reporting the best outcome
of different iterations of cross-validation. Moreover, there is a lack
of discussion on why a particular ML model is utilized. Recently, there
is a trend towards using more complex ML models such as random forest,
extreme gradient boosting and neural networks without a discussion on if
and how much model interpretibility is necessary for the study (11--14).
The lack of transparency on modeling methodology and model selection
negatively impact model reproducibility and reliability. We need to
strive toward better machine learning practices by (1) implementing
consistent and correct machine learning pipelines; (2) selecting ML
models that reflect the goal of the study as it will inform our
expectations of model accuracy, complexity, interpretibility and
computational efficiency.

To showcase reliable ML methodologies and to shed light on how much ML
model selection can affect modeling results, we performed an empirical
analysis comparing several different ML models using the same dataset
and a robust ML pipeline. We used a previously published colorectal
cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
patients. We built ML models using fecal 16S rRNA gene sequences to
predict patients with normal colons or patients with colonic tumors
which are called screen relevant neoplasias (SRN). The study had 261
normal and 229 SRN samples. We established modeling pipelines for
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost which are ML models that increase in
complexity respectively. Our ML pipeline utilized held-out test data to
evaluate generalization and prediction performance of each ML model. The
mean test AUROC varied from 0.598 (std ± 0.044) to 0.697 (std ± 0.037).
Random forest had the highest mean AUROC for detecting SRN. Despite the
simplicity, the L2-regularized logistic regression followed random
forest in performance. In terms of computational efficiency, L2 logistic
regression trained the fastest (0.202 hours, std ± 0.019), while XGBoost
took the longest (162.843 hours, std ± 10.018). We found that mean
cross-validation and testing AUROC varied only 0.01, which highlights
the importance of a seperate held-out test set and consistent
preprocessing of the data prior to evaluation. Aside from evaluating
generalization and classification performances for each of these models,
this study established standards for modeling pipelines of
microbiome-associated machine learning models.

@@

The data has 6920 features (6920 OTUs) and a two-class label that
defines the colonic health of the patients (SRN or normal). We
established modeling pipelines for binary classification with
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost to emphasize the differences in model
accuracy, complexity, interpretibility and computational efficiency due
to model selection. We randomly split the data into stratified
training/validation and test sets 100 times with a 80/20 proportion
{[}Figure 1{]}. For each data-split, training/validation set consisted
of 393 patients (209 SRN), while the test set was composed of 97
patients (52 SRN). The training/validation data was used for training
purposes and validation of hyperparameter selection, and the test set
was used for evaluation purposes. Validation of hyperparameter selection
was performed by randomly splitting the training/validation set into
stratified training and validation sets 100 times with a 80/20
proportion {[}Figure 1{]}. For L2-regularized logistic regression, L1
and L2 support vector machines (SVM) with linear and radial basis
function kernels we tuned the \textbf{cost} hyperparameter which
determines the regularization strength where smaller values specify
stronger regularization. For SVM with radial basis function kernel we
also tuned \textbf{sigma} hyperparameter which determines the reach of a
single training instance where for a high value of sigma, the SVM
decision boundary will be dependent on the points that are closest to
the decision boundary. For the decision tree model, we tuned the
\textbf{depth of the tree} where deeper the tree, the more splits it
has. For random forest, we tuned the \textbf{number of features} to
consider when looking for the best tree split. For xgboost, we tuned for
\textbf{learning rate} and the \textbf{fraction of samples} to be used
for fitting the individual base learners. We validated the prediction
performances of each hyperparameter setting (Figure S1{]}.

@@

We evaluated the prediction performance of seven binary classification
models when applied to held-out test data over 100 data-splits using
AUROC as classification performance metric. Random Forest had the
highest mean AUROC for detecting SRN, 0.697 (std ± 0.037) {[}Figure
2{]}. L2 logistic regression and XGBoost had significantly lower AUROC
values, 0.676 (std ± 0.042) and 0.678 (std ± 0.04) {[}Figure 2{]}.
Random forest also had significanly higher performances than L2 linear
SVM, RBF SVM and decision tree which had mean AUROC values of 0.671 (std
± 0.045), 0.663 (std ± 0.044), and 0.598 (std ± 0.044), respectively
{[}Figure 2{]}. We also evaluated the generalization performance of each
classifier by comparing their mean cross-validation AUROC and mean
testing AUROC. We found that mean cross-validation and testing AUROC
difference for each model was below 0.01.

@@

We interpreted the feature weights of L1 and L2 SVM with linear kernel
and regression coefficients of L2 logistic regression using the training
data. To get a sense of the most important features in linear models, we
calculated the mean weights of all the features over the 100 data-splits
for each model. {[}Figure 4{]}. In the three linear models, FIT was the
feature that drove the detection of SRNs, followed by eight OTUs which
belong to family \emph{Lachnospiraceae} and \emph{Ruminococcaceae} (four
with positive signs; Otu01239, Otu00742, Otu00008, Otu00659, and four
with negative signs; Otu00015, Otu000150, Otu00609, Otu00629). To get
the importance of features in a decision tree, we used gini importance
which allows us to use the improvement in the split-criterion is the
importance measure attributed to the splitting variable

@@

Linear models trained faster than non-linear models. L2 logistic
Rregression and L1 and L2 SVM with linear kernel had training times of
13.2 min, (std ± 1.8), 563.4 min, (std ± 46.8) and 12.6 min, (std ± 1.8)
respectively. Whereas, SVM with radial basis function kernel, decision
tree, random forest and xgboost had training times of 101.32 hrs, (std ±
10.02), 162.84 hrs, (std ± 3.99), 0.2 hrs, (std ± 0.02) and 64.71 hrs,
(std ± 9.85), respectively {[}Figure 4{]}.

@@

The classification models were built to classify normal versus adenoma,
normal versus carcinoma, and high versus low SCFA concentrations. The
regression models were built to classify the SCFA concentrations of
acetate, butyrate, and propionate regardless of disease status. The
featThe bacterial abundances are the features used to predict colonic
disease status. Bacterial abundances are discrete data in the form of
Operational Taxonomic Unit (OTU) counts. There are 6920 OTUs for each
sample. FIT levels are continuous data present for each sample. Because
the data are in different scales, features are transformd by scaling
each feature to a {[}0-1{]} range

To evaluate the model performances, we randomly split the data into
training/validation and test sets 100 times with a 80/20 proportion.
Data-splits were stratified for diagnosis or SCFA concentration levels
in classification models. For each data-split, training/validation set
consisted of x patients (x carcinoma, y adenoma, z normal), while the
test set was composed of x patients (x carcinoma, y adenoma, z normal).
The training/validation data was used for training purposes and
validation of hyperparameter selection, and the test set was used for
evaluation purposes. Validation of hyperparameter selection was
performed by randomly splitting the training/validation set into
stratified training and validation sets 100 times with a 80/20
proportion {[}Figure 1{]}.

\paragraph{Data collection}\label{data-collection}

The data used for this analysis are stool bacterial abundances, stool
hemoglobin levels and clinical information of the patients recruited by
Great Lakes-New England Early Detection Research Network study. These
data were obtained from Sze et al (15). The stool samples were provided
by recruited adult participants who were undergoing scheduled screening
or surveillance colonoscopy. Colonoscopies were performed and fecal
samples were collected from participants in four locations: Toronto (ON,
Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA).
Patients' colonic disease status was defined by colonoscopy with
adequate preparation and tissue histopathology of all resected lesions.
Patients with an adenoma greater than 1 cm, more than three adenomas of
any size, or an adenoma with villous histology were classified as
advanced adenoma. Study had 172 patients with normal colonoscopies, 198
with adenomas and 120 with carcinomas. Of the 198 adenomas, 109 were
identified as advanced adenomas. Stool provided by the patients was used
for Fecal Immunological Tests (FIT) which measure human hemoglobin
concentrations and for 16S rRNA gene sequencing to measure bacterial
population abundances. The bacterial abundance data was generated by Sze
et al, by processing 16S rRNA sequences in Mothur (v1.39.3) using the
default quality filtering methods, identifying and removing chimeric
sequences using VSEARCH and assigning to OTUs at 97\% similarity using
the OptiClust algorithm (16--18).

@@

The colonic disease status is re-defined as two encompassing classes;
Normal or Screen Relevant Neoplasias (SRNs). Normal class includes
patients with non-advanced adenomas or normal colons whereas SRN class
includes patients with advanced adenomas or carcinomas. Colonic disease
status is the label predicted with each classifier. The bacterial
abundances and FIT results are the features used to predict colonic
disease status. Bacterial abundances are discrete data in the form of
Operational Taxonomic Unit (OTU) counts. There are 6920 OTUs for each
sample. FIT levels are continuous data present for each sample. Because
the data are in different scales, features are transformd by scaling
each feature to a {[}0-1{]} range (Table 1).

\paragraph{Learning the Classifier}\label{learning-the-classifier}

To train and validate our model, labeled data is randomly split 80/20
into a training set and testing set. Then, seven binary class
classifiers, L2 logistic regression, L1 and L2 linear suppor vector
machines (SVM), radial basis function SVM, decision tree, random forest
and XGBoost, are learned. The training set is used for training purposes
and validation of hyperparameter selection, and the test set is used for
evaluation purposes. Hyperparameters are selected using 5-fold
cross-validation with 100-repeats on the training set. Since the colonic
disease status are not uniformly represented in the data, 5-fold splits
are stratified to maintain the overall label distribution on the
training set.

\paragraph{Classifier Performance}\label{classifier-performance}

The classification performance of learned classifier is evaluated on the
labeled held-out testing set. The optimal classifier with optimal
hyperparameters selected in the cross-validation step is used to produce
a prediction for the testing set. The performance of this prediction is
measured in terms of the sensitivity and specificity, in addition to
Area Under the Curve (AUC) metrics. This process of splitting the data,
learning a classifier with cross-validation, and testing the classifier
is repeated on 100 different splits. In the end cross-validation AUC and
testing AUC averaged over the 100 different training/test splits are
reported. Hyperparameter budget and performance for each split is also
reported.

>>>>

Advances in sequencing technology and decreasing costs of generating 16S
rRNA gene sequences have allowed rapid exploration of human associated
microbiome and its health implications. Currently, the human microbiome
field is growing at an unprecedented rate and as a result, there is an
increasing demand for methods that identify associations between members
of the microbiome and human health. However, this is difficult as human
associated microbial communities are remarkably complex and uneven. It
is unlikely that a single species can explain a disease. Instead,
subsets of those communities, in relation to one another and to their
host, account for the differences in health outcomes. Machine learning
(ML) methods are effective at recognizing and highlighting patterns in
complex microbial datasets. Therefore, researchers have started to
explore the utility of ML models that use microbiota associated
biomarkers to predict human health and to understand the microbial
ecology of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases (IBD), obesity, and type 2 diabetes (1--11).
However, currently the field's use of ML lacks clarity and consistency
on which methods are used and how these methods are implemented. Most
notably, there are misleading practices of using ``leaky'' ML pipelines
where models are tested on training data or reporting the best outcome
of different iterations of cross-validation. Moreover, there is a lack
of discussion on why a particular ML model is utilized. Recently, there
is a trend towards using more complex ML models such as random forest,
extreme gradient boosting and neural networks without a discussion on if
and how much model interpretibility is necessary for the study (11--14).
The lack of transparency on modeling methodology and model selection
negatively impact model reproducibility and reliability. We need to
strive toward better machine learning practices by (1) implementing
consistent and reliable machine learning pipelines; (2) selecting ML
models that reflect the goal of the study as it will inform our
expectations of model accuracy, complexity, interpretibility and
computational efficiency.

To showcase reliable ML pipeline and to shed light on how much ML model
selection can affect modeling results, we performed an empirical
analysis comparing several different ML models using the same dataset
and with a robust ML pipeline. We used a previously published colorectal
cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
patients. We built ML models using fecal 16S rRNA gene sequences to
predict patients with normal colons or patients with colonic tumors
which are called screen relevant neoplasias (SRN). The study had 261
normal and 229 SRN samples. We established modeling pipelines for
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost. Our ML pipeline utilized held-out test data
to evaluate generalization and prediction performance of each ML model.
The mean test AUROC varied from 0.598 (std ± 0.044) to 0.697 (std ±
0.037). Random forest had the highest mean AUROC for detecting SRN.
Despite the simplicity, the L2-regularized logistic regression followed
random forest in performance. In terms of computational efficiency, L2
logistic regression trained the fastest (0.202 hours, std ± 0.019),
while XGBoost took the longest (162.843 hours, std ± 3.986). We found
that mean cross-validation and testing AUROC varied only 0.01, which
highlights the importance of a seperate held-out test set and consistent
preprocessing of the data prior to evaluation. Aside from evaluating
generalization and classification performances for each of these models,
this study established standards for modeling pipelines of
microbiome-associated machine learning models.

@@

The data has 6920 features (6920 OTUs) and a two-class label that
defines the colonic health of the patients (SRN or normal). All the
cases were independently labeled through colonoscopies. We established
modeling pipelines for a binary prediction task with L2-regularized
logistic regression, L1 and L2 support vector machines (SVM) with linear
and radial basis function kernels, a decision tree, random forest and
XGBoost to emphasize the differences in model accuracy, complexity,
interpretibility and computational efficiency due to model selection. We
randomly split the data into training/validation and test sets 100 times
with a 80/20 proportion {[}Figure 1{]}. Since the colonic health is not
uniformly represented in the data, data-splits are stratified to
maintain the overall label distribution on the training set. For each
data-split, training/validation set consisted of 393 patients (209 SRN),
while the test set was composed of 97 patients (52 SRN). The
training/validation data was used for training purposes and validation
of hyperparameter selection, and the test set was used for evaluation
purposes. Validation of hyperparameter selection was performed using 100
randomizations of five-fold cross-validation on the training/validation
set {[}Figure 1{]}. Similar to the initial data-split, five-fold
cross-validation was also stratified to maintain the overall label
distribution on the training and validation sets. We validated the
cross-validation performances of each hyperparameter setting over the
100 randomizations and selected the best performing hyper-parameter
setting to train the full training/validation dataset {[}Figures S1 and
S2{]}. We then used the held-out test set to evaluate the prediction
performance of each ML model.

@@

We evaluated the prediction performance of seven binary classification
models when applied to held-out test data over 100 data-splits using the
area under the receiver operating characteristic curve (AUROC) as the
discriminative performance metric. Random Forest had the highest mean
AUROC for detecting SRN, 0.697 (std ± 0.037) {[}Figure 2{]}. L2 logistic
regression and XGBoost had significantly lower AUROC values, 0.676 (std
± 0.042) and 0.678 (std ± 0.04) {[}Figure 2{]}. Random forest also had
significanly higher performances than L2 linear SVM, RBF SVM and
decision tree which had mean AUROC values of 0.671 (std ± 0.045), 0.663
(std ± 0.044), and 0.598 (std ± 0.044), respectively {[}Figure 2{]}. We
also evaluated the generalization performance of each classifier by
comparing their mean cross-validation AUROC and mean testing AUROC. We
found that mean cross-validation and testing AUROC difference for each
model was below 0.01.

@@

We interpreted the feature weights of L1 and L2 SVM with linear kernel
and regression coefficients of L2 logistic regression using the training
data. We calculated the mean weights of all the features fromt the 100
randomizations of five-fold cross-validation {[}Figure 3a{]}. In the
three linear models, OTUs that drove the detection of SRNs belong to
family \emph{Lachnospiraceae} and \emph{Ruminococcaceae} (OTUs x, y, z).
To explain the importance of features in non-linear models, we used
permutation importance on the held-out test data {[}Figure 3b{]}.
Perfectly correlated OTUs were grouped and permuted together to get a
better representation of their effect in the prediction performance of
the model. Non-correlated OTUs were permuted individually.

@@

Linear models trained faster than non-linear models. L2 logistic
Rregression and L1 and L2 SVM with linear kernel had training times of
0.2 hours, (std ± 0.02), 0.2 hours, (std ± 0.03), and 0.2 hours, (std ±
0.03), respectively. Whereas, SVM with radial basis function kernel,
decision tree, random forest and xgboost had training times of 9.4
hours, (std ± 0.8), 64.7 hours, (std ± 9.9), 101.3 hours, (std ± 10) and
162.8 hours, (std ± 4), respectively {[}Figure 4{]}.

@@

\paragraph{Data collection and study
population}\label{data-collection-and-study-population}

The data used for this analysis are stool bacterial abundances and
clinical information of the patients recruited by Great Lakes-New
England Early Detection Research Network study. These data were obtained
from Sze et al (15). The stool samples were provided by recruited adult
participants who were undergoing scheduled screening or surveillance
colonoscopy. Colonoscopies were performed and fecal samples were
collected from participants in four locations: Toronto (ON, Canada),
Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA). Patients'
colonic disease status was defined by colonoscopy with adequate
preparation and tissue histopathology of all resected lesions. Patients
with an adenoma greater than 1 cm, more than three adenomas of any size,
or an adenoma with villous histology were classified as advanced
adenoma. Study had 172 patients with normal colonoscopies, 198 with
adenomas and 120 with carcinomas. Of the 198 adenomas, 109 were
identified as advanced adenomas. Stool provided by the patients was used
for 16S rRNA gene sequencing to measure bacterial population abundances.
The bacterial abundance data was generated by Sze et al, by processing
16S rRNA sequences in Mothur (v1.39.3) using the default quality
filtering methods, identifying and removing chimeric sequences using
VSEARCH and assigning to OTUs at 97\% similarity using the OptiClust
algorithm (16--18).

@@

The colonic health of the patient was defined as two encompassing
classes; Normal or Screen Relevant Neoplasias (SRNs). Normal class
includes patients with non-advanced adenomas or normal colons whereas
SRN class includes patients with advanced adenomas or carcinomas. The
bacterial abundances are the features used to predict colonic health of
the patients. Bacterial abundances are discrete data in the form of
Operational Taxonomic Unit (OTU) counts. OTU counts were set to the size
of our smallest sample and were subsampled at the same distances. They
were then transformd by scaling to a {[}0-1{]} range.

\paragraph{Model training and
evaluation}\label{model-training-and-evaluation}

For L2-regularized logistic regression, L1 and L2 support vector
machines (SVM) with linear and radial basis function kernels we tuned
the \textbf{cost} hyperparameter which determines the regularization
strength where smaller values specify stronger regularization. For SVM
with radial basis function kernel we also tuned \textbf{sigma}
hyperparameter which determines the reach of a single training instance
where for a high value of sigma, the SVM decision boundary will be
dependent on the points that are closest to the decision boundary. For
the decision tree model, we tuned the \textbf{depth of the tree} where
deeper the tree, the more splits it has. For random forest, we tuned the
\textbf{number of features} to consider when looking for the best tree
split. For xgboost, we tuned for \textbf{learning rate} and the
\textbf{fraction of samples} to be used for fitting the individual base
learners. Models were trained using the machine learning wrapper caret
package (v.6.0.81) in R (v.3.5.0).

\textbf{Statistical analysis workflow.} Data summaries, statistical
analysis, and data visualizations were performed using R (v.3.5.0) with
the tidyverse package (v.1.2.1). We compared the AUROC values of the
seven ML models by Wilcoxon rank sum tests to determine the best
discriminative performance.

\textbf{Code availability.} The code for all sequence curation and
analysis steps including an Rmarkdown version of this manuscript is
available at
\url{https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/}.
